Importance of using the correct background list for over-representation analysis

### 2023-07-09 manuscript/manuscript.Rmd
intro work

---
title: "Importance of using the correct background list for over-representation analysis"
author: "Mark Ziemann and Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

@@

[Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)

@@

>>>>

---
title: "Importance of using the correct background list for over-representation analysis"
author: "Mark Ziemann, Barry Schroeter, Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

@@

[Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
[Barry Schroeter<sup>1</sup>]()
[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)

@@

The fold enrichment score (ES) is commonly used to express the degree of enrichment and is a ratio-of-ratios.
The foreground ratio is the number of selected genes belonging a functional class (A) divided by the
number of detected genes in the functional class (B).
The background ratio is the number of selected genes not belonging to the functional class (C) divided
by the number of detected genes not in the functional class (D).

$$
ES=\frac{ A / B}{ C / D}
$$

Both equations illustrate that correct definition of the number of background genes is important to
obtain accurate enrichment scores and significance values.
A common error in ORA of differential gene expression is to use all genes in the genome as the
background @Wijesooriya2022-li.
This leads to a severe over-estimation of the number of background genes as there are ~45k annotated
genes in the human genome (according to Ensembl v109), but RNA-seq robustly detects only ~15k genes.
This leads to inflation of p-values and results that share only ~40% similarity with properly conducted
ORA analysis.
This issue is even more severe for other omics, where the number of detected genes is lower, for example
proteomics and single cell RNA-seq where there are only a few thousand detected analytes.
To address this issue, most major ORA tools allow for a user-provided background list of detected genes.

While testing some of these tools, we observed an unusual behaviour whereby the number of background
genes reported by the tool varied significantly from the length of the gene list we provided.
Moreover, this number also varied when we used a different library of gene function annotations.
Puzzled by this behaviour, we browsed the computational biology forums and were met with other confused
users wondering why the number of genes in the background list would vary like this.
The tool developers kindly clarified the situation, by stating that the tool removed genes from the
entire analysis if they were not members of a functional category.
We thought this behaviour odd, because it has the potential to significantly alter the test being
conducted, away from it's original application.

In this paper, we examine the effect of this filtering step on ORA of real RNA-seq data, use simulated
RNA-seq data to determine the impact upon precision and recall, and provide practical recommendations
to end users to obtain the most accurate results.
